### 2023-02-03 inputs/data/data_import.qmd
Initial commit

```{r}
library(opendatatoronto)
library(tidyverse)
library(dplyr)

#raw_response_data <-
#  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
#  filter(name == "Response Time (2013 - 2021)") |>
#  get_resource()

#write_csv(
#  x = raw_response_data$`Land Ambulance Services`,
#  file = "raw_response_data.csv"
#)

#raw_emergent_care_data <-
#  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
#  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
#  get_resource() |>
#  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")

#write_csv(
#  x = raw_emergent_care_data,
#  file = "raw_emergent_care_data.csv"
#)

#raw_response_numbered_data <-
#  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
#  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
#  get_resource() |>
#  filter(`Pre-Hospital Emergency Care Type` == "Arrival to Life Threatening Calls 90% of Time")

#write_csv(
#  x = raw_response_numbered_data,
#  file = "raw_response_numbered_data.csv",
#)

raw_incident_data <-
  list_package_resources("c21f3bd1-e016-4469-abf5-c58bb8e8b5ce") |>
  filter(name == "paramedic-services-incident-data-2017-2021") |>
  get_resource()

write_csv(
  x = raw_incident_data$`2021`,
  file = "raw_incident_data.csv",
)
```

### 2023-02-03 outputs/paper/analysis.qmd
Initial commit

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that the Toronto's paramedics meet and often exceed response time standards, and provided stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---

## Introduction

A city's emergency medical services (EMS) are integral to its health care system. The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance. These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires. Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour]. This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].

In Toronto, inhabitants receive emergency medical assistance through the Toronto Paramedic Services. As of 2020, the Toronto Paramedic Services employed 1,186 paramedics operating 234 ambulances out of 45 stations across the city [@cite_toronto_paramedic_2020_report]. The service provides all land ambulatory responses within the city, including on-scene treatment and stabilisation of patients, transport to a hospital, and transport between hospitals. The Province of Ontario and the City of Toronto set specific requirements for the length of time an ambulance should take when responding to an emergency call, which can be anywhere between 6 to 25 minutes. While the service strives to achieve these mandated goals, they face staffing, budget, and other institutional challenges that impact its ability to provide care. Recently, the service gained attention for slower response times because of staffing shortages in hospitals and in the service itself [@cite_paramedic_shortages]. These staffing shortages decrease the number of available paramedics while simultaneously increasing the time it takes for a patient to enter a hospital, tying up paramedics at hospitals. The recent COVID-19 pandemic also impacted the service, with most paramedics stating that morale is "either worse or as bad as it has ever been" [@cite_paramedic_morale].

Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival. This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments. Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service. \newpage

## Paramedic Response Data

```{r}
#| messages: false
#| echo: false
#| warning: false
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv")) |>
  clean_names()
raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"))
raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv")) |>
  clean_names()
```

### Scope of Responsibility

Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by the graph below (@fig-1). This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
#| label: fig-1
raw_emergent_care_data |>
  # Define plot 
  ggplot(aes(
    x = Year, 
    y = Actual
  )) +

  # Define axis labels
  labs(y = "Emergency Responses", x = "Year") +

  # Set the theme
  theme_minimal() +

  # Set axis formats
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_x_continuous(breaks=seq(2014,2021,1)) +

  # Render bar chart
  geom_bar(stat="identity", fill = "deepskyblue3")

```

\newpage

To further illustrate the responsibility of Toronto's paramedics, the table below (@fig-2) outlines the number of incidents, and ambulance dispatches the service fielded in 2021. The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires. These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day. Most calls and incidents need between 1 and three ambulances. However, some severe incidents placed an even greater strain on the service's resources:

-   74 fire emergencies required 5 or more ambulances, with one incident requiring a total of 22 ambulances.
-   46 medical emergencies required 5 or more ambulances, with one incident requiring a total of 11 ambulances.
-   33 motor vehicle accidents required 5 or more ambulances, with one incident requiring a total of 11 ambulances.

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Emergency Medical Incidents and Dispatched Ambulances (in 2021)"
#| label: fig-2
clean_incident_data <- raw_incident_data |>
  # Group data by incident type
  group_by(incident_type) |>

  # Select only incident type and arrived units
  select(incident_type, units_arrived_at_scene) |>

  # Add new columns summarizing data
  summarize(
    # Calculate total number of incident per incident type
    total_incidents = table(incident_type),

    # Calculate total number of ambulances responding to incidents
    units_sum = sum(units_arrived_at_scene, na.rm = TRUE),

    # Calculate the mean of ambulances per incident 
    the_mean = signif(mean(units_arrived_at_scene, na.rm = TRUE), digits = 2),

    # Calculate the mean of ambulances per incident 
    dispatches_day = round(sum(units_arrived_at_scene, na.rm = TRUE)/365, digits = 0), 
  ) |>

  # Rename column headers
  rename("Incident Type" = "incident_type", "Incidents" = "total_incidents", "Dispatches" = "units_sum", "Per Incident" = "the_mean", "Per Day" = "dispatches_day")

# Rename "-" in data to more verbose label
clean_incident_data[1, "Incident Type"] = "Other Incident"

# Render table
clean_incident_data |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold=TRUE
  ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE) |>
  column_spec(1, width = "13em") 
```

\newpage

### Mandated Response Times

In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries.

Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*. This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively (source; source). CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service. The table below (@fig-3) outlines the City of Toronto's response times for each CTAS level.

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Mandated Response Times for the Toronto Paramedic Services (as of 2021)"
#| label: fig-3
raw_response_data |>
  # Select type of response and estimated response plan 
  select(patient_category, plan_in_minutes) |>

  # Remove identical rows
  distinct() |>

  # Create new descriptor column 
  cbind(descriptor = c("Sudden Cardiac Arrest", "Resuscitative care", "Emergent care", "Urgent care", "Less urgent care", "Non-urgent care")) |>

  # Rename all columns to human-friendly format
  rename("Response Level" = "patient_category", "Response Time (Mins)" = "plan_in_minutes", "Level Descriptor" = "descriptor") |>

  # Render table
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE)

```

It is important to note that not all calls are mandated to receive a response time within these parameters. Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times. In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame. However, prior to 2016, only 60% of SCA calls required a response within the mandated time. \newpage

### Actual Response Times

The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of Calls Meeting Mandated Response Times (since 2013)"
#| label: fig-4
raw_response_data |>
  # Select desired variables
  select(patient_category, performance_in_percentage, plan_in_percentage, response_time_standard_year) |>

  # Remove rows with no data
  filter(performance_in_percentage != "NA") |>

  # Plot graphs, define x axis
  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Calls Meeting Response Time", x = "Year") +

  # Set the theme
  theme_minimal() +

  # Rotate X-axis labels
  theme(axis.text.x = element_text(angle=70, vjust=0.7)) +

  # Set axis formats
  scale_x_continuous(breaks=seq(2013,2021,2)) +

  # Render lines
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue") +

  # Merge graphs into one figure
  facet_wrap(~patient_category)

```

There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan. In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan. The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes. \newpage

## Discussion

The data used for analysis in this paper was collected and published by the Toronto Paramedic Service. As such, the data is open to institutional bias. The Toronto Paramedic Services determines what data is collected and the methods used for collection. It also has specific goals for data collection: in the case of response times, they must collect the data so they can report their performance to Ontario's Ministry of Health. These goals can impact the data and cause the service to omit significant data points, such as the lack of a measured, case-by-case response time. The data also has other limitations, such as the lack of metrics after 2021.

When analysing a paramedic service's efficacy, an important consideration is whether response times based on the "golden hour" are valuable metrics. In principle, the sooner a paramedic can respond to a call leads to greater efficiency, as it shortens the time paramedics spend on a call and helps patients receive more immediate care. However, healthcare professionals are debating the relevance of the "golden hour" as the foundation of emergency medical services. One study found that "there was no association between EMS intervals and mortality," suggesting that the "golden hour," in the context of EMS, did not impact a patient's chances of survival [@cite_study_1]. A separate study focusing on rural areas found that longer transport times to a trauma center "did not significantly influence survival" [@cite_study_2]. Contrasting these findings is a study from Quebec, which found that spending over 60 minutes before arriving at a trauma center had a "statistically significant" impact on a patient's odds of survival, although this study only surveyed severely injured patients requiring advanced life support (ALS) [@cite_study_3]. These studies reveal that response times based on the "golden hour" principle are of little help in determining a paramedic service's efficacy, except for the most urgent cases that require advanced measures.

With this new limited scope in mind and focusing on only urgent cases, data from the Toronto Paramedic Services reveals that their quick response times effectively improved the chances of survival for critical patients. Patients in Toronto suffering from a stroke face a mortality rate of 11.3%, below the province's average of 12.9% [@cite_stroke_mortality]. Patients in Toronto suffering from a STEMI, or a heart attack caused by a complete blockage of the coronary artery, face a mortality rate of 10.9%, slightly lower than the province's average of 11.2% [@cite_stemi_mortality]. It is important to note that the Toronto Paramedic Service did not include this data in the packages made available on Open Data Toronto.

The Toronto Paramedic Services should include more comprehensive metrics in the data sets they provide to Open Data Toronto. As revealed in this paper, the service's current data sets lack vital information on case-by-case response times and patient outcomes. Where permitted, the service should also include the type of medical emergency addressed by paramedics. The inclusion of these expanded data points would allow for a more meaningful analysis of their data. \newpage

## Inclusive Acknowledgements

**Land Acknowledgement**

We wish to acknowledge the land on which this data was collected and analysed. For thousands of years, it has been the traditional land of the Huron-Wendat, the Seneca, and the Mississaugas of the Credit. Today, this meeting place is still the home to many Indigenous people from across Turtle Island and we are grateful to have the opportunity to work on this land.

This statement was originally prepared in consultation with First Nations House and the Elders Circle for use at the University of Toronto [@cite_land_acknowledgement].\

**Data Acknowledgement**

The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto]. This paper uses three data sets from Open Data Toronto, each provided by the Toronto Paramedic Services:

-   Paramedic Services Incident Data[^1]
-   Land Ambulance Response Time Standard[^2]
-   Pre-Hospital Emergency Care Performance Metrics[^3]\

This project contains information licensed under the Open Government Licence – Toronto^[Open Government License – Toronto: https://open.toronto.ca/open-data-license/].

[^1]: Paramedic Services Incident Data available at: https://open.toronto.ca/dataset/paramedic-services-incident-data

[^2]: Land Ambulance Response Time Standard available at: https://open.toronto.ca/dataset/land-ambulance-response-time-standard

[^3]: Pre-Hospital Emergency Care Performance Metrics available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics

**Resources Acknowledgment**

The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r]. The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse]. \newpage

## References

<!-- Span required to show references as a numbered section -->

### 2023-02-03 outputs/paper/copy.qmd
Initial commit

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez"
abstract: "We consider data about the motor vehicle collisions in Toronto." 
date: 2023-02-03
editor: visual
# toc: TRUE
bibliography: references.bib
---

## Introduction
Emergency medical technicians (EMTs) and paramedics provide pre-hospital care to those in need of urgent medical assistance. These professionals respond to every call from the public for aid, and are capable of stabalizing patients suffering from serious illness and trauma. Emergency medical services operate on the principle of the “golden hour,” which is the period of time between an initial injury and treatment when the patient has the highest odds of survival (source). This “golden hour” is the foundation of all medical response efforts, and typically requires ambulances arrive on the scene of an emergency in under 10 minutes (source).

In Toronto, the public receives emergency medical aid through the Toronto Paramedic Services. As of 2020, the Toronto Paramedic Services employed 1,186 paramedics who operate 234 ambulances out of 45 stations across the city (2020 annual report). The service provides all land ambulatory responses within the city, including paramedicine treatment and transportation to and between hospitals. Provincial and municipal legislation requires that the paramedic service provides a response time of _____. While the service strives to meed these mandated goals, they face staffing, budget, and institutional challenges that often impact their ability to provide care. The service gained public attention in recent years due to slower response times because hospitals are unable to efficently process new patients, leaving paramedics to wait with their patients until they can be seen by a doctor (). Due to the COVID-19 pandemic, a majority of paramedics in the service feel that morale is “either worse or as bad as it has ever been,” leaving many in search of new employment ().

Despite the stresses placed on Toronto’s paramedics, they must respond to emergencies quickly so they can provide patients with the best odds of survival. This paper investigates the response time of paramedics to see if the service is maximixing their patients’ chances of survival. In particular, this paper reviews historical response times to uncover trends and

## Paramedic Response Data
```{r}
#| messages: false
#| echo: false
#| warning: false

library(janitor)
raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv")) |>
  clean_names()
raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"))
#raw_response_numbered_data <- readr::read_csv(here::here("inputs/data/raw_response_numbered_data.csv"))
#raw_response_numbered_data
```
### Source and Data Collection

We examined motor vehicle collisions (MVC) in Toronto using data from Toronto Police Services via Open Data Toronto [@cite_opendatatoronto]. Built with the R programming language [@cite_r] and ggplot2 [@cite_ggplot]

### Historic Trends

| Response Level^[Based on the Canadian Triage Acuity Scale (CTAS)] | Target Response Time | Level Descriptor |
|--------|:------|-------:|
| SCA | 6 mins | Sudden Cardiac Arrest |
| CTAS Level 1 | 8 mins | Resuscitative care |
| CTAS Level 2 | 10 mins | Emergent care |
| CTAS Level 3 | 15 mins | Urgent care |
| CTAS Level 4 | 20 mins | Less urgent care |
| CTAS Level 5 | 25 mins | Non-urgent care |
https://www.health.gov.on.ca/en/pro/programs/emergency_health/land/default.aspx#CTAS

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of SCA Cases Meeting Response Time"
#| label: fig-1
library(lubridate)
library(tidyverse)
library(dplyr)

raw_response_data |> 
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA")
```

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of SCA Cases Meeting Response Time"
#| label: fig-1
library(lubridate)
library(tidyverse)
library(dplyr)

raw_response_data |>
#  filter(patient_category == "SCA") |>

  select(patient_category, performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +

  scale_x_continuous(breaks=seq(2013,2021,2)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue") +
  facet_wrap(~patient_category)

```

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
#| label: fig-7
library(lubridate)
library(tidyverse)
library(dplyr)
library(scales)
library(RColorBrewer)

raw_emergent_care_data |>
  ggplot(aes(
    x = Year, 
    y = Actual
  )) +

  # Define axis labels
  labs(y = "Emergency Responses", x = "Year") +

  # Set the theme
  theme_minimal() +
  scale_color_brewer(palette="Blues") +

  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_x_continuous(breaks=seq(2014,2021,1)) +

  # Render
  geom_bar(stat="identity")

```

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of SCA Cases Meeting Response Time"
#| label: fig-1
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "SCA") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of CTAS Level 1 Cases Meeting Response Time"
#| label: fig-2
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "CTAS 1") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of CTAS Level 2 Cases Meeting Response Time"
#| label: fig-3
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "CTAS 2") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of CTAS Level 3 Cases Meeting Response Time"
#| label: fig-4
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "CTAS 3") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of CTAS Level 4 Cases Meeting Response Time"
#| label: fig-5
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "CTAS 4") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```
```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Percent of CTAS Level 5 Cases Meeting Response Time"
#| label: fig-6
library(lubridate)
library(tidyverse)
library(dplyr)
raw_response_data |>
  filter(patient_category == "CTAS 5") |>
  select(performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
  filter(performance_in_percentage != "NA") |>

  ggplot(aes(x=response_time_standard_year)) +

  # Define axis labels
  labs(y = "% of Cases Meeting Response Time", x = "Year") +

  # geom_hline(yintercept=0.75, linetype="dashed", color = "red") +

  # Set the theme
  theme_minimal() +

  scale_x_continuous(breaks=seq(2013,2021,1)) +

  # Render
  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
  geom_line(aes(y = performance_in_percentage), color="steelblue")
```

### 2021 Responses

## Discussion
- Response times measured by the Toronto Paramedic Service, open to institutional bias
- Limitations in the dataset only show years up to 2021

## References

### 2023-02-03 inputs/data/00_raw_data_import.R

@@ -1,0 +1,23 @@
+library(opendatatoronto)
+library(tidyverse)
+library(dplyr)
+
+raw_response_data <-
+  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
+  filter(name == "Response Time (2013 - 2021)") |>
+  get_resource()
+
+write_csv(
+  x = raw_response_data$`Land Ambulance Services`,
+  file = "raw_response_data"
+)
+
+raw_emergent_care_data <-
+  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
+  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
+  get_resource()
+
+write_csv(
+  x = raw_emergent_care_data$`Land Ambulance Services`,
+  file = "raw_emergent_care_data"
+)

### 2023-02-03 inputs/data/data_import.qmd
Add tests and simulation, improve code comments, remove extraneous files

@@
 ```{r}
+# Define required libraries
 library(opendatatoronto)
@@
 
-#raw_response_data <-
-#  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
-#  filter(name == "Response Time (2013 - 2021)") |>
-#  get_resource()
+# Fetch Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
+raw_response_data <-
+  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
+  filter(name == "Response Time (2013 - 2021)") |>
+  get_resource()
 
-#write_csv(
-#  x = raw_response_data$`Land Ambulance Services`,
-#  file = "raw_response_data.csv"
-#)
+# Write data to a .csv file
+write_csv(
+  x = raw_response_data$`Land Ambulance Services`,
+  file = "raw_response_data.csv"
+)
 
-#raw_emergent_care_data <-
-#  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
-#  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
-#  get_resource() |>
-#  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")
+# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+raw_emergent_care_data <-
+  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
+  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
+  get_resource() |>
+  filter(`Pre-Hospital Emergency Care Type` == "Emergency Request for Service")
 
-#write_csv(
-#  x = raw_emergent_care_data,
-#  file = "raw_emergent_care_data.csv"
-#)
+# Write data to a .csv file
+write_csv(
+  x = raw_emergent_care_data,
+  file = "raw_emergent_care_data.csv"
+)
 
-#raw_response_numbered_data <-
-#  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
-#  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
-#  get_resource() |>
-#  filter(`Pre-Hospital Emergency Care Type` == "Arrival to Life Threatening Calls 90% of Time")
-
-#write_csv(
-#  x = raw_response_numbered_data,
-#  file = "raw_response_numbered_data.csv",
-#)
-
+# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
 raw_incident_data <-
@@
 
+# Write data to a .csv file
 write_csv(

### 2023-02-03 outputs/paper/analysis.qmd
Add tests and simulation, improve code comments, remove extraneous files

@@
 #| warning: false
+# Define required libraries 
 library(lubridate)
@@
 library(kableExtra)
-raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv")) |>
+library(here)
+library(readr)
+
+# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
+raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
   clean_names()
-raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"))
-raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv")) |>
+
+# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE)
+
+# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
+raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
   clean_names()
@@
 -   Land Ambulance Response Time Standard[^2]
--   Pre-Hospital Emergency Care Performance Metrics[^3]\
+-   Pre-Hospital Emergency Care Performance Metrics[^3]
 
-This project contains information licensed under the Open Government Licence – Toronto^[Open Government License – Toronto: https://open.toronto.ca/open-data-license/].
+This project contains information licensed under the Open Government Licence -- Toronto[^4].\
 
-[^1]: Paramedic Services Incident Data available at: https://open.toronto.ca/dataset/paramedic-services-incident-data
+[^1]: Paramedic Services Incident Data is available at: https://open.toronto.ca/dataset/paramedic-services-incident-data
 
-[^2]: Land Ambulance Response Time Standard available at: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
+[^2]: Land Ambulance Response Time Standard is available at: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
 
-[^3]: Pre-Hospital Emergency Care Performance Metrics available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+[^3]: Pre-Hospital Emergency Care Performance Metrics is available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+
+[^4]: Open Government License -- Toronto is available at: https://open.toronto.ca/open-data-license/
 
@@
 The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r].
-The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse].
+The paper also uses a number of R packages, including: dplyr [@cite_tool_dplyr], ggplot2 [@cite_tool_ggplot2], here [@cite_tool_here], janitor [@cite_tool_janitor], kableExtra [@cite_tool_kableextra], knitr [@cite_tool_knitr], lubridate [@cite_tool_lubridate], opendatatoronto [@cite_tool_opendatatoronto], readr [@cite_tool_readr], RColorBrewer [@cite_tool_rcolorbrewer], scales [@cite_tool_scales], and tidyverse [@cite_tool_tidyverse].
 \newpage
@@
 ## References
-
 <!-- Span required to show references as a numbered section -->

### 2023-02-03 tests.qmd
Add tests and simulation, improve code comments, remove extraneous files

<!--
#### Preamble ####
# Purpose: Testing Data from the Toronto Paramedic Service
# Prerequisites: Need to have downloaded the data using [inputs/data/data_import.qmd]
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "Testing Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Workspace set-up ####
library(tidyverse)
library(janitor)
library(here)
library(readr)
```

```{r}
# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
  clean_names() |>
  filter(performance_in_percentage != "NA")

# Test that the target performance is less than or equal to 75%
raw_response_data$plan_in_percentage |> max() <= 0.75

# Test that the reported year is greater than or equal to 2013
raw_response_data$response_time_standard_year |> min() >= 2013

# Test that the actual performance data is less than or equal to 100%
raw_response_data$performance_in_percentage |> max() <= 1.00

# Test that the actual performance data is greater than or equal to 0%
raw_response_data$performance_in_percentage |> min() >= 0.00

```

```{r}
# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE) |>
  clean_names()

# Test that the reported year is greater than or equal to 2014
raw_emergent_care_data$year |> min() >= 2014

# Test that the actual responses is a number
raw_emergent_care_data$actual |> class() == "numeric"

# Test that the actual responses is not greater than Toronto's population
raw_emergent_care_data$actual |> max() <= 3000000

# Test that the actual responses is not 0 
raw_emergent_care_data$actual |> min() >= 0
```

```{r}
# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
  clean_names()

raw_incident_data

# Test that the units arrived on scene is not greater than the city's number of paramedics
raw_incident_data$units_arrived_at_scene |> max() <= 1186

# Test that the units arrive on scene is greater than zero
raw_incident_data$units_arrived_at_scene |> min() >= 0

# Test that the dispatch time is a date 
raw_incident_data$dispatch_time |> class() == "POSIXct"

# Test that the incident type contains is comprised of characters
raw_incident_data$incident_type |> class() == "character"

```

### 2023-02-03 toronto_paramedic_response.qmd
Add tests and simulation, improve code comments, remove extraneous files

<!--
#### Preamble ####
# Purpose: Read in data from the Toronto Paramedic Service
# Prerequisites: Need to know where to get Toronto Paramedic Services data
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "Simulating Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Workspace set-up ####
# install.packages("tidyverse") # Do this once per computer
# install.packages("janitor") # This also

library(tidyverse) # A collection of data-related packages
library(janitor) # Helps clean datasets
```

```{r}
simulated_response_data <-
  tibble(
    # Patient categories, predefined and set terms
    "Patient Category" = c(
      "SCA", 
      "CTAS 1",
      "CTAS 2",
      "CTAS 3",
      "CTAS 4",
      "CTAS 5"
    ),
    # Response time mandates, predefined and set numbers
    "Plan in Minutes" = c(
      6, 
      8,
      10,
      15,
      20,
      25
    ),
    # Performance mandates, predefined and set percentage
    "Plan in Percentage" = sample(
      x = 0.75,
      size = 6,
      replace = TRUE
    ),
    # Actual percentage of response times meeting performance plan, varying between 60% and 99%
    "Performance in Percentage" = sample(
      seq(0.6, 0.99, by = 0.001),
      size = 6,
      replace = TRUE
    ),
    # Sample year associated with data
    "Response Time Standard Year" = c(
      2021, 
      2021,
      2021,
      2021,
      2021,
      2021
    ),
)

simulated_response_data
```

```{r}
simulated_emergent_care_data <-
  tibble(
    # Define the type of pre-hospital care: while the raw data has multiple values for this, we only use Emergency Requests for Service
    "Pre-Hospital Emergency Care Type" = sample(
      x = "Emergency Request for Service",
      size = 8,
      replace = TRUE
    ),

    # List all years between 2014 and 2021, inclusive
    "Year" = c(2014:2021),

    # Vary the data between 200,000 and 300,000
    "Actual" = sample(
      x = c(200000:350000),
      size = 8,
      replace = TRUE
    )
)

simulated_emergent_care_data
```

```{r}
simulated_incident_data <-
  tibble(
    # Vary the date between January 2, 2021 and December 31, 2021
    "Dispatch_Time" = sample(
      seq(as.Date('2021/01/01'), as.Date('2021/12/31'), by="day"), 
      size = 1000,
      replace = TRUE
    ),

    # Randomly pick an incident type1000 times
    "Incident_Type" = sample(
      x = c(
        "Medical",
        "Motor Vehicle Accident",
        "Emergency Transfer",
        "Fire",
        "Airport Standby",
        "-"
      ),
      size = 1000,
      replace = TRUE
    ),

    # Randomly select the number of units arriving on scene, between 1 and 22
    "Units_Arrived_At_Scene" = sample(
      x = c(1:22),
      size = 1000,
      replace = TRUE
    )
  )

simulated_incident_data
```

### 2023-02-03 inputs/data/00_raw_data_import.R

@@ -1,23 +1,0 @@
-library(opendatatoronto)
-library(tidyverse)
-library(dplyr)
-
-raw_response_data <-
-  list_package_resources("8b068db1-95b8-4ea9-b942-8a7ce3a5fbd6") |>
-  filter(name == "Response Time (2013 - 2021)") |>
-  get_resource()
-
-write_csv(
-  x = raw_response_data$`Land Ambulance Services`,
-  file = "raw_response_data"
-)
-
-raw_emergent_care_data <-
-  list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
-  filter(name == "Pre-hospital emergency care performance metrics 2014-2021") |>
-  get_resource()
-
-write_csv(
-  x = raw_emergent_care_data$`Land Ambulance Services`,
-  file = "raw_emergent_care_data"
-)

### 2023-02-03 data-testing.qmd
Update comments and descriptions; add .pdf files to gitignore

--- 2023-02-03 tests.qmd
@@
 raw_response_data$performance_in_percentage |> min() >= 0.00
-
 ```
@@
 raw_incident_data$incident_type |> class() == "character"
-
 ```

### 2023-02-03 inputs/data/data_import.qmd
Update comments and descriptions; add .pdf files to gitignore

@@
+<!--
+#### Preamble ####
+# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
+# Author: Sebastian Rodriguez
+# Email: me@srod.ca
+# Date: 3 February 2023
+# GitHub: https://github.com/seb646/toronto-paramedic-responses
+-->
+---
+title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
+format: html
+author: "Sebastian Rodriguez"
+---
 ```{r}
-# Define required libraries
+#### Workspace set-up ####
 library(opendatatoronto)
@@
 library(dplyr)
+```
 
+```{r}
 # Fetch Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
@@
 )
+```
 
+```{r}
 # Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
@@
 )
+```
 
+```{r}
 # Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data

### 2023-02-03 outputs/paper/analysis.qmd
Update comments and descriptions; add .pdf files to gitignore

@@
+<!--
+#### Preamble ####
+# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
+# Author: Sebastian Rodriguez
+# Email: me@srod.ca
+# Date: 3 February 2023
+# GitHub: https://github.com/seb646/toronto-paramedic-responses
+-->
 ---
@@
 ---
-
-## Introduction
-
-A city's emergency medical services (EMS) are integral to its health care system.
-The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance.
-These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires.
-Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour].
-This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].
-
-In Toronto, inhabitants receive emergency medical assistance through the Toronto Paramedic Services.
-As of 2020, the Toronto Paramedic Services employed 1,186 paramedics operating 234 ambulances out of 45 stations across the city [@cite_toronto_paramedic_2020_report].
-The service provides all land ambulatory responses within the city, including on-scene treatment and stabilisation of patients, transport to a hospital, and transport between hospitals.
-The Province of Ontario and the City of Toronto set specific requirements for the length of time an ambulance should take when responding to an emergency call, which can be anywhere between 6 to 25 minutes.
-While the service strives to achieve these mandated goals, they face staffing, budget, and other institutional challenges that impact its ability to provide care.
-Recently, the service gained attention for slower response times because of staffing shortages in hospitals and in the service itself [@cite_paramedic_shortages].
-These staffing shortages decrease the number of available paramedics while simultaneously increasing the time it takes for a patient to enter a hospital, tying up paramedics at hospitals.
-The recent COVID-19 pandemic also impacted the service, with most paramedics stating that morale is "either worse or as bad as it has ever been" [@cite_paramedic_morale].
-
-Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival.
-This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments.
-Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.
-\newpage
-
-## Paramedic Response Data
-
 ```{r}
@@
 #| warning: false
+#### Workspace set-up ####
 # Define required libraries 
@@
 
+## Introduction
+A city's emergency medical services (EMS) are integral to its health care system.
+The paramedics and emergency medical technicians working for the service provide pre-hospital assessment, stabilisation, and care to people needing medical assistance.
+These professionals respond to every call from the public for aid, from broken bones and heart attacks to car accidents and house fires.
+Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour].
+This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].
+
+In Toronto, inhabitants receive emergency medical assistance through the Toronto Paramedic Services.
+As of 2020, the Toronto Paramedic Services employed 1,186 paramedics operating 234 ambulances out of 45 stations across the city [@cite_toronto_paramedic_2020_report].
+The service provides all land ambulatory responses within the city, including on-scene treatment and stabilisation of patients, transport to a hospital, and transport between hospitals.
+The Province of Ontario and the City of Toronto set specific requirements for the length of time an ambulance should take when responding to an emergency call, which can be anywhere between 6 to 25 minutes.
+While the service strives to achieve these mandated goals, they face staffing, budget, and other institutional challenges that impact its ability to provide care.
+Recently, the service gained attention for slower response times because of staffing shortages in hospitals and in the service itself [@cite_paramedic_shortages].
+These staffing shortages decrease the number of available paramedics while simultaneously increasing the time it takes for a patient to enter a hospital, tying up paramedics at hospitals.
+The recent COVID-19 pandemic also impacted the service, with most paramedics stating that morale is "either worse or as bad as it has ever been" [@cite_paramedic_morale].
+
+Despite the challenges that Toronto's paramedics face, they must respond to emergencies quickly to provide patients with the best chances of survival.
+This paper investigates the response times of the Toronto Paramedic Service compared to the mandated requirements set by the local and provincial governments.
+Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.
+\newpage
+
+## Paramedic Response Data
 ### Scope of Responsibility
-
 Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility.
@@
 ```
-
 \newpage
@@
 ### Mandated Response Times
-
 In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times.
@@
 ### Actual Response Times
-
 The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times.
@@
 ## Discussion
-
 The data used for analysis in this paper was collected and published by the Toronto Paramedic Service.
@@
 **Land Acknowledgement**
-
 We wish to acknowledge the land on which this data was collected and analysed.
@@
 **Data Acknowledgement**
-
 The data used in this paper was obtained through Open Data Toronto, an initiative to provide openly accessible data about the city [@cite_about_opendatatoronto].
@@
 **Resources Acknowledgment**
-
 The primary tool used to analyse data in this paper is R, an open-source statistical programming language [@cite_tool_r].
@@
 ## References
-<!-- Span required to show references as a numbered section -->
+<!-- Comment required to show references as a numbered section -->

### 2023-02-03 data-simulation.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

--- 2023-02-03 toronto_paramedic_response.qmd
@@
   tibble(
-    # Define the type of pre-hospital care: while the raw data has multiple values for this, we only use Emergency Requests for Service
+    # Define the type of pre-hospital care (we only use this one value)
     "Pre-Hospital Emergency Care Type" = sample(
@@
 
-    # Vary the data between 200,000 and 300,000
+    # Randomly select numbers between 200,000 and 300,000
     "Actual" = sample(
@@
   tibble(
-    # Vary the date between January 2, 2021 and December 31, 2021
+    # Randomly pick dates between January 2, 2021 and December 31, 2021
     "Dispatch_Time" = sample(
@@
 
-    # Randomly pick an incident type1000 times
+    # Randomly pick an incident type 1000 times
     "Incident_Type" = sample(

### 2023-02-03 inputs/data/data_import.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

@@
-<!--
+---
+title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
+format: html
+author: "Sebastian Rodriguez"
+---
+```{r}
 #### Preamble ####
@@
 # GitHub: https://github.com/seb646/toronto-paramedic-responses
--->
----
-title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
-format: html
-author: "Sebastian Rodriguez"
----
+```
+
 ```{r}

### 2023-02-03 outputs/paper/analysis.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

--- 2023-02-03 outputs/paper/analysis.qmd
@@
 This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments.
-It reveals that the Toronto's paramedics meet and often exceed response time standards, and provided stroke and STEMI heart attack patients with a chance of survival greater than the provincial average."
+It reveals that Toronto's paramedics meet and often exceed response time standards, and provide stroke and STEMI heart attack patients with a chance of survival greater than the provincial average."
 date: today
@@
 ---
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#### Preamble ####
+# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
+# Author: Sebastian Rodriguez
+# Email: me@srod.ca
+# Date: 3 February 2023
+# GitHub: https://github.com/seb646/toronto-paramedic-responses
+```
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#### Workspace set-up ####
+# Define required libraries 
+library(lubridate)
+library(tidyverse)
+library(dplyr)
+library(knitr)
+library(janitor)
+library(scales)
+library(RColorBrewer)
+library(ggplot2)
+library(kableExtra)
+library(here)
+library(readr)
+```
+
+```{r}
+#| messages: false
+#| echo: false
+#| warning: false
+#### Fetch and clean relevant data sets ####
+# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
+raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
+  clean_names()
+
+# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE) |>
+  clean_names()
+
+# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
+raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
+  clean_names()
+```
 
@@
 
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-# Define required libraries 
-library(lubridate)
-library(tidyverse)
-library(dplyr)
-library(knitr)
-library(janitor)
-library(scales)
-library(RColorBrewer)
-library(ggplot2)
-library(kableExtra)
-library(here)
-library(readr)
-
-# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
-raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
-  clean_names()
-
-# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
-raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE)
-
-# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
-raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
-  clean_names()
-```
-
 ### Scope of Responsibility
@@
 As the city's population grows, so too does the need for medical assistance.
-There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by the graph below (@fig-1).
+There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by @fig-1 [@cite_emergency_care_data].
 This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system.
@@
   ggplot(aes(
-    x = Year, 
-    y = Actual
+    x = year, 
+    y = actual
   )) +
@@
 ```
-
 \newpage
 
-To further illustrate the responsibility of Toronto's paramedics, the table below (@fig-2) outlines the number of incidents, and ambulance dispatches the service fielded in 2021.
+To further illustrate the responsibility of Toronto's paramedics, @fig-2 outlines the number of incidents, and ambulance dispatches the service fielded in 2021 [@cite_paramedic_incident_data].
 The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires.
 These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day.
-Most calls and incidents need between 1 and three ambulances.
+Most calls and incidents need between 1 and three ambulances [@cite_paramedic_incident_data].
 However, some severe incidents placed an even greater strain on the service's resources:
@@
 The *Ambulance Act* mandates varying response times based on the severity of a patient's injury.
-To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage Acuity Scale (CTAS).
-CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries.
+To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage & Acuity Scale (CTAS).
+CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries [@cite_ctas].
 
 Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*.
-This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively (source; source).
+This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively  [@cite_ambulance_act].
 CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service.
-The table below (@fig-3) outlines the City of Toronto's response times for each CTAS level.
+@fig-3 outlines the City of Toronto's response times for each CTAS level [@cite_ambulance_response_data].
 
@@
 It is important to note that not all calls are mandated to receive a response time within these parameters.
-Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times.
+Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times [@cite_ambulance_act].
 In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame.
-However, prior to 2016, only 60% of SCA calls required a response within the mandated time.
+However, prior to 2016, only 60% of SCA calls required a response within the mandated time [@cite_ambulance_response_data].
 \newpage
@@
 The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times.
-The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan.
+The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan [@cite_ambulance_response_data].
 To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].
@@
 There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan.
-In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan.
+In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan [@cite_ambulance_response_data].
 The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes.

### 2023-02-03 scripts/01-data_import.qmd
Organized scripts into a folder

--- 2023-02-03 inputs/data/data_import.qmd
@@
   x = raw_response_data$`Land Ambulance Services`,
-  file = "raw_response_data.csv"
+  file = "../inputs/data/raw_response_data.csv"
 )
@@
   x = raw_emergent_care_data,
-  file = "raw_emergent_care_data.csv"
+  file = "../inputs/data/raw_emergent_care_data.csv"
 )
@@
   x = raw_incident_data$`2021`,
-  file = "raw_incident_data.csv",
+  file = "../inputs/data/raw_incident_data.csv",
 )

### 2023-02-03 scripts/02-data_testing.qmd
Organized scripts into a folder

--- 2023-02-03 data-testing.qmd
@@
-<!--
+---
+title: "Testing Data from the Toronto Paramedic Service"
+format: html
+author: "Sebastian Rodriguez"
+---
+```{r}
 #### Preamble ####
@@
 # GitHub: https://github.com/seb646/toronto-paramedic-responses
--->
----
-title: "Testing Data from the Toronto Paramedic Service"
-format: html
-author: "Sebastian Rodriguez"
----
+```
+
 ```{r}
@@
 
-raw_incident_data
-
 # Test that the units arrived on scene is not greater than the city's number of paramedics

### 2023-02-03 outputs/paper/paper.qmd
Changed analysis.qmd to paper.qmd

--- 2023-02-03 outputs/paper/analysis.qmd
@@
 ---
+
 ```{r}
@@
 As the city's population grows, so too does the need for medical assistance.
-There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by @fig-1 [@cite_emergency_care_data].
+There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by [@fig-1; @cite_emergency_care_data].
 This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system.
@@
 ```
+
 \newpage
@@
 
-This project contains information licensed under the Open Government Licence -- Toronto[^4].\
-
 [^1]: Paramedic Services Incident Data is available at: https://open.toronto.ca/dataset/paramedic-services-incident-data
@@
 [^3]: Pre-Hospital Emergency Care Performance Metrics is available at: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+
+This project contains information licensed under the Open Government Licence -- Toronto[^4].\
 
@@
 ## References
+
 <!-- Comment required to show references as a numbered section -->

### 2023-02-03 outputs/paper/paper.qmd
General updates and bug fixes
* Moved data cleaning to a separate script file
* Renamed emergent care data set to a more meaningful title: performance data
* Updated readme.md

@@
 #| warning: false
+#| output: false
 #### Workspace set-up ####
@@
 #### Fetch and clean relevant data sets ####
-# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
-raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
-  clean_names()
+# Fetch the the clean response data
+clean_response_data <- readr::read_csv(here::here("inputs/data/clean_response_data.csv"), show_col_types = FALSE)
 
-# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
-raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE) |>
-  clean_names()
+# Fetch clean performance data
+clean_performance_data <- readr::read_csv(here::here("inputs/data/clean_performance_data.csv"), show_col_types = FALSE)
 
-# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
-raw_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
-  clean_names()
+# Fetch the clean incident response
+clean_incident_data <- readr::read_csv(here::here("inputs/data/clean_incident_data.csv"), show_col_types = FALSE)
 ```
@@
 #| label: fig-1
-raw_emergent_care_data |>
+clean_performance_data |>
   # Define plot 
@@
     x = year, 
-    y = actual
+    y = ems_responses
   )) +
@@
 #| label: fig-2
-clean_incident_data <- raw_incident_data |>
+summarized_incident_data <- clean_incident_data |>
   # Group data by incident type
@@
 
-# Rename "-" in data to more verbose label
-clean_incident_data[1, "Incident Type"] = "Other Incident"
+# Rename "-" in data to a more meaningful label
+summarized_incident_data[1, "Incident Type"] = "Other Incident"
+
+# Order rows alphabetically by incident type
+summarized_incident_data <- summarized_incident_data[order(summarized_incident_data$`Incident Type`),]
 
 # Render table
-clean_incident_data |>
+summarized_incident_data |>
+  # Table configuration 
   knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
+
+  # Add headers for table data
   add_header_above(
@@
     align = "c",
-    bold=TRUE
+    bold = TRUE
   ) |>
+
+  # Style the table 
   kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
+
+  # Make the table headers bold
   row_spec(0,bold=TRUE) |>
+
+  # Increase width of first column
   column_spec(1, width = "13em") 
@@
 #| label: fig-3
-raw_response_data |>
+clean_response_data |>
   # Select type of response and estimated response plan 
@@
 #| label: fig-4
-raw_response_data |>
-  # Select desired variables
-  select(patient_category, performance_in_percentage, plan_in_percentage, response_time_standard_year) |>
-
-  # Remove rows with no data
-  filter(performance_in_percentage != "NA") |>
-
+clean_response_data |>
   # Plot graphs, define x axis
@@
 
+  guides(shape=TRUE) +
+
   # Merge graphs into one figure
   facet_wrap(~patient_category) 
-
 ```

### 2023-02-03 scripts/01-data_import.qmd
General updates and bug fixes
* Moved data cleaning to a separate script file
* Renamed emergent care data set to a more meaningful title: performance data
* Updated readme.md

@@
 ```{r}
+#| warning: false
 # Fetch Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
@@
 ```{r}
-# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
-raw_emergent_care_data <-
+#| warning: false
+# Fetch "Pre-Hospital Emergency Care Performance Metrics" from Open Data Toronto
+# Link: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
+raw_performance_data <-
   list_package_resources("174ace54-bcc7-4885-85c1-76094ab4a0ca") |>
@@
 write_csv(
-  x = raw_emergent_care_data,
-  file = "../inputs/data/raw_emergent_care_data.csv"
+  x = raw_performance_data,
+  file = "../inputs/data/raw_performance_data.csv"
 )
@@
 ```{r}
-# Fetch Open Data Toronto's "Paramedic Services Incident Data" package: https://open.toronto.ca/dataset/paramedic-services-incident-data
+#| warning: false
+# Fetch Paramedic Services Incident Data from Open Data Toronto
+# Link: https://open.toronto.ca/dataset/paramedic-services-incident-data
 raw_incident_data <-
@@
 )
+
+# Note: creating this .csv may take some time as it's a large data set
 ```

### 2023-02-03 scripts/02-data_cleaning.qmd
General updates and bug fixes
* Moved data cleaning to a separate script file
* Renamed emergent care data set to a more meaningful title: performance data
* Updated readme.md

---
title: "Import raw csv data, clean it, and generate new csv files"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Import raw csv data, clean it, and generate new csv files
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
```

```{r}
#### Workspace set-up ####
library(tidyverse)
library(dplyr)
library(here)
library(readr)
```

```{r}
# Fetch and clean the raw response data
clean_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |>

  # Select the desired columns
  select(patient_category, performance_in_percentage, plan_in_minutes, plan_in_percentage, response_time_standard_year) |>

  # Remove rows with no data
  filter(performance_in_percentage != "NA")

# Create a new cleaned csv file
write_csv(
  x = clean_response_data,
  file = "../inputs/data/clean_response_data.csv",
)
```

```{r}
# Fetch and clean the raw performance data
clean_performance_data <- readr::read_csv(here::here("inputs/data/raw_performance_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |>

  # Select the desired columns
  select(year, actual) |>

  # Rename column to meaninful label
  rename("ems_responses" = "actual")

# Create a new cleaned csv file
write_csv(
  x = clean_performance_data,
  file = "../inputs/data/clean_performance_data.csv",
)
```

```{r}
# Fetch and clean the raw incident data
clean_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>
  # Change column naming format
  clean_names() |>

  # Select the desired columns
  select(dispatch_time, incident_type, units_arrived_at_scene)

# Create a clean csv file
write_csv(
  x = clean_incident_data,
  file = "../inputs/data/clean_incident_data.csv",
)
```

### 2023-02-03 scripts/03-data_testing.qmd
General updates and bug fixes
* Moved data cleaning to a separate script file
* Renamed emergent care data set to a more meaningful title: performance data
* Updated readme.md

---
title: "Testing Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Testing Data from the Toronto Paramedic Service
# Prerequisites: Need to have downloaded the data using [inputs/data/data_import.qmd]
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
```

```{r}
#### Workspace set-up ####
library(tidyverse)
library(janitor)
library(here)
library(readr)
```

```{r}
# Fetch the clean response data 
clean_response_data <- readr::read_csv(here::here("inputs/data/clean_response_data.csv"), show_col_types = FALSE)

# Test that the target performance is less than or equal to 75%
clean_response_data$plan_in_percentage |> max() <= 0.75

# Test that the reported year is greater than or equal to 2013
clean_response_data$response_time_standard_year |> min() >= 2013

# Test that the actual performance data is less than or equal to 100%
clean_response_data$performance_in_percentage |> max() <= 1.00

# Test that the actual performance data is greater than or equal to 0%
clean_response_data$performance_in_percentage |> min() >= 0.00
```

```{r}
# Fetch the clean performance data
clean_performance_data <- readr::read_csv(here::here("inputs/data/clean_performance_data.csv"), show_col_types = FALSE)

# Test that the reported year is greater than or equal to 2014
clean_performance_data$year |> min() >= 2014

# Test that EMS responses is a number
clean_performance_data$ems_responses |> class() == "numeric"

# Test that EMS responses is not greater than Toronto's population
clean_performance_data$ems_responses |> max() <= 3000000

# Test that EMS responses is not 0 
clean_performance_data$ems_responses |> min() >= 0
```

```{r}
# Fetch the clean incident data
clean_incident_data <- readr::read_csv(here::here("inputs/data/clean_incident_data.csv"), show_col_types = FALSE)

# Test that the units arrived on scene is not greater than the city's number of paramedics
clean_incident_data$units_arrived_at_scene |> max() <= 1186

# Test that the units arrive on scene is greater than zero
clean_incident_data$units_arrived_at_scene |> min() >= 0

# Test that the dispatch time is a date 
clean_incident_data$dispatch_time |> class() == "POSIXct"

# Test that the incident type contains is comprised of characters
clean_incident_data$incident_type |> class() == "character"
```

### 2023-02-03 scripts/04-data_simulation.qmd
General updates and bug fixes
* Moved data cleaning to a separate script file
* Renamed emergent care data set to a more meaningful title: performance data
* Updated readme.md

<!--
#### Preamble ####
# Purpose: Read in data from the Toronto Paramedic Service
# Prerequisites: Need to know where to get Toronto Paramedic Services data
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "Simulating Data from the Toronto Paramedic Service"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Workspace set-up ####
# install.packages("tidyverse") # Do this once per computer
# install.packages("janitor") # This also

library(tidyverse) # A collection of data-related packages
library(janitor) # Helps clean datasets
```

```{r}
# Simulate data based on Land Ambulance Response Time Standard
# Link: https://open.toronto.ca/dataset/land-ambulance-response-time-standard 
simulated_response_data <-
  tibble(
    # Patient categories, predefined and set terms
    "patient_category" = c(
      "SCA", 
      "CTAS 1",
      "CTAS 2",
      "CTAS 3",
      "CTAS 4",
      "CTAS 5"
    ),
    # Response time mandates, predefined and set numbers
    "plan_in_minutes" = c(
      6, 
      8,
      10,
      15,
      20,
      25
    ),
    # Performance mandates, predefined and set percentage
    "plan_in_percentage" = sample(
      x = 0.75,
      size = 6,
      replace = TRUE
    ),
    # Actual percentage of response times meeting performance plan, varying between 60% and 99%
    "performance_in_percentage" = sample(
      seq(0.6, 0.99, by = 0.001),
      size = 6,
      replace = TRUE
    ),
    # Sample year associated with data
    "response_time_standard_tear" = c(
      2021, 
      2021,
      2021,
      2021,
      2021,
      2021
    ),
)

simulated_response_data
```

```{r}
# Simulate data based on Pre-Hospital Emergency Care Performance Metrics 
# Link: https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
simulated_performance_data <-
  tibble(
    # List all years between 2014 and 2021, inclusive
    "year" = c(2014:2021),

    # Randomly select numbers between 200,000 and 300,000
    "ems_responses" = sample(
      x = c(200000:350000),
      size = 8,
      replace = TRUE
    )
)

simulated_performance_data
```

```{r}
# Simulate data based on Paramedic Services Incident Data
# Link: https://open.toronto.ca/dataset/paramedic-services-incident-data
simulated_incident_data <-
  tibble(
    # Randomly pick dates between January 2, 2021 and December 31, 2021
    "dispatch_time" = sample(
      seq(as.Date('2021/01/01'), as.Date('2021/12/31'), by="day"), 
      size = 1000,
      replace = TRUE
    ),

    # Randomly pick an incident type 1000 times
    "incident_type" = sample(
      x = c(
        "Medical",
        "Motor Vehicle Accident",
        "Emergency Transfer",
        "Fire",
        "Airport Standby",
        "Other"
      ),
      size = 1000,
      replace = TRUE
    ),

    # Randomly select the number of units arriving on scene, between 1 and 22
    "units_arrived_at_scene" = sample(
      x = c(1:22),
      size = 1000,
      replace = TRUE
    )
  )

simulated_incident_data
```

### 2023-02-03 outputs/paper/paper.qmd
Update paper.qmd
* Added additional comments
* Fixed code styling

@@
   # Make the table headers bold
-  row_spec(0,bold=TRUE) |>
+  row_spec(0, bold = TRUE) |>
 
@@
   # Create new descriptor column 
-  cbind(descriptor = c("Sudden Cardiac Arrest", "Resuscitative care", "Emergent care", "Urgent care", "Less urgent care", "Non-urgent care")) |>
+  cbind(descriptor = c(
+    "Sudden Cardiac Arrest", 
+    "Resuscitative care", 
+    "Emergent care", 
+    "Urgent care", 
+    "Less urgent care", 
+    "Non-urgent care")
+  ) |>
 
   # Rename all columns to human-friendly format
-  rename("Response Level" = "patient_category", "Response Time (Mins)" = "plan_in_minutes", "Level Descriptor" = "descriptor") |>
+  rename(
+    "Response Level" = "patient_category", 
+    "Response Time (Mins)" = "plan_in_minutes", 
+    "Level Descriptor" = "descriptor"
+  ) |>
 
@@
   knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
+
+  # Style the table 
   kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
-  row_spec(0,bold=TRUE)
+
+  # Make the table headers bold
+  row_spec(0, bold = TRUE)
 

### 2023-02-03 scripts/03-data_testing.qmd
General cleaning and adding comments
* Added a note to the incident data testing to describe an issue where the test returns two Booleans, one of which is false. Clarified this incident for the user.
* Fixed minor spacing issue with paper.qmd

@@
 clean_incident_data$incident_type |> class() == "character"
+
+# Note: due to an issue with the dispatch_time's column formatting as a date,
+# the second returned Boolean may be false. As long as the first value returns 
+# true, the data passes the test.
 ```

### 2023-02-03 outputs/paper/paper.qmd
General cleaning and adding comments
* Added a note to the incident data testing to describe an issue where the test returns two Booleans, one of which is false. Clarified this incident for the user.
* Fixed minor spacing issue with paper.qmd
Added more code comments

@@
 #| warning: false
+#| output: false
 #### Preamble ####
@@
 # GitHub: https://github.com/seb646/toronto-paramedic-responses
-```
 
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-#| output: false
 #### Workspace set-up ####
@@
 library(readr)
-```
 
-```{r}
-#| messages: false
-#| echo: false
-#| warning: false
-#### Fetch and clean relevant data sets ####
 # Fetch the the clean response data
@@
 #| label: fig-1
+#### Plot graph on the number of EMS responses in Toronto ####
 clean_performance_data |>
@@
   geom_bar(stat="identity", fill = "deepskyblue3")
-
 ```
@@
 #| label: fig-2
+#### Create a table summarizing emergency medical incident data in Toronto ####
 summarized_incident_data <- clean_incident_data |>
@@
 #| label: fig-3
+#### Create a table outlining CTAS levels and response times ####
 clean_response_data |>
@@
 #| label: fig-4
+#### Create graphs plotting mandated and actual response times ####
 clean_response_data |>

### 2023-02-03 scripts/02-data_cleaning.qmd
Added more code comments

@@
 # GitHub: https://github.com/seb646/toronto-paramedic-responses
-```
 
-```{r}
 #### Workspace set-up ####
@@
 ```{r}
-# Fetch and clean the raw response data
+#### Fetch and clean the raw response data ####
 clean_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
@@
 ```{r}
-# Fetch and clean the raw performance data
+ #### Fetch and clean the raw performance data  ####
 clean_performance_data <- readr::read_csv(here::here("inputs/data/raw_performance_data.csv"), show_col_types = FALSE) |>
@@
 ```{r}
-# Fetch and clean the raw incident data
+#### Fetch and clean the raw incident data ####
 clean_incident_data <- readr::read_csv(here::here("inputs/data/raw_incident_data.csv"), show_col_types = FALSE) |>

### 2023-02-03 outputs/paper/paper.qmd
Paper updates
* Included a summary of findings in the introduction
* Fixed formatting issues

@@
 Current emergency medical services operate on the principle of the "golden hour," which is the 60 minutes between an initial injury and arrival at a trauma center when a patient has the highest statistical chance of survival [@cite_golden_hour].
-This "golden hour" is the foundation of all emergency medical responses and typically requires medical personnel to arrive on the scene of an injury in under 10 minutes [@cite_golden_hour_10_mins].
+This "golden hour" is the foundation of all emergency medical responses and requires medical personnel to arrive on scene in under 10 minutes [@cite_golden_hour_10_mins].
 
@@
 Specifically, the paper aims to place response times within a larger context of the service's responsibility and interrogate the relevance of the "golden hour" principle to determine the efficacy of the service.
+
+The findings indicated a growing trend in the number of emergency medical calls, an increase of approximately 10,000 per year since 2014 [@cite_emergency_care_data].
+Despite this increase, the Toronto Paramedic Services met and often exceeded the response times set by both the Province of Ontario and the City of Toronto.
+The paper also revealed a significant lack of paramedic metrics from the data sets available on Open Data Toronto.
+Without metrics on case-by-case response times and patient outcomes, other sources were required to inform data analysis.
+Overall, the response time of Toronto's paramedics contributed to stroke and mortality rates lower than Ontario's average, giving the city's inhabitants greater chances of survival when suffering from life-threatening injuries.
 \newpage
@@
 Where permitted, the service should also include the type of medical emergency addressed by paramedics.
-The inclusion of these expanded data points would allow for a more meaningful analysis of their data.
+The inclusion of these expanded data points would allow for a more meaningful analysis.
 \newpage

### 2023-02-04 outputs/paper/paper.qmd
Added legend to graphs

@@
   # Define axis labels
-  labs(y = "Emergency Responses", x = "Year") +
+  labs(y = "Emergency Medical Responses", x = "Year") +
 
@@
 
+  # Define styles
+  theme(
+    axis.title.x = element_text(size = 9, family = "serif", face = "bold"),
+    axis.text.x = element_text(size = 8, family = "serif"),
+    axis.title.y = element_text(size = 9, family = "serif", face = "bold"),
+    axis.text.y = element_text(size = 8, family = "serif")
+  ) +
+
   # Set axis formats
   scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
-  scale_x_continuous(breaks=seq(2014,2021,1)) +
+  scale_x_continuous(breaks = seq(2014, 2021, 1)) +
 
@@
 #| warning: false
+#| label: fig-4
 #| fig-cap: "Percent of Calls Meeting Mandated Response Times (since 2013)"
-#| label: fig-4
+#| fig-align: center
+#| fig-width: 10
+#| fig-height: 5
 #### Create graphs plotting mandated and actual response times ####
+colors <- c("Actual Response" = "steelblue", "Mandated Response" = "darkred")
+
 clean_response_data |>
   # Plot graphs, define x axis
-  ggplot(aes(x=response_time_standard_year)) +
+  ggplot(aes(x = response_time_standard_year)) +
 
   # Define axis labels
-  labs(y = "% of Calls Meeting Response Time", x = "Year") +
+  labs(
+    y = "Percent of Calls", 
+    x = "Year",
+    color = "Legend"
+  ) +
 
@@
   # Rotate X-axis labels
-  theme(axis.text.x = element_text(angle=70, vjust=0.7)) +
+  theme(
+    axis.title.x = element_text(size = 15, family = "serif", face = "bold"),
+    axis.text.x = element_text(size = 14, family = "serif", angle = 70, vjust = 0.7),
+    axis.title.y = element_text(size = 15, family = "serif", face = "bold"),
+    axis.text.y = element_text(size = 14, family = "serif"),
+    legend.title = element_text(size = 15, family = "serif", face = "bold"),
+    legend.text = element_text(size = 14, family = "serif"),
+    strip.text.x = element_text(size = 15, family = "serif", face = "bold"),
+    strip.text.y = element_text(size = 14, family = "serif", face = "bold")
+  ) +
 
   # Set axis formats
-  scale_x_continuous(breaks=seq(2013,2021,2)) +
+  scale_x_continuous(breaks=seq(2013, 2021, 2)) +
 
   # Render lines
-  geom_line(aes(y = plan_in_percentage), color = "darkred") + 
-  geom_line(aes(y = performance_in_percentage), color="steelblue") +
+  geom_line(aes(y = plan_in_percentage, color = "Mandated Response")) + 
+  geom_line(aes(y = performance_in_percentage, color = "Actual Response")) +
 
-  guides(shape=TRUE) +
+  scale_color_manual(values = colors) +
 
